---
title: |
  <h1 style='font-size:38px; font-weight:bold;'>Exploring how dietary patterns shape the global obesity crisis</h1><br>
  <h3 style='font-style:italic; font-size:22px;'>How our diets shape our destiny: A data driven journey into global obesity</h3><br>
  <p style='font-size:20px;'><strong>Course Name & Code:</strong> Data Visualisation and Communication (<strong>MATH2270</strong>)</p>

author: "Prathibha Magesh"
output: slidy_presentation

---


# Introduction

## A Global Weight Gain Story

In the last few decades, our plates and waistlines have changed dramatically.

Once, traditional home-cooked meals nourished families. Today, processed snacks and fast food dominate diets across both developed and developing nations. With drive-thrus replacing dinner tables, the world is gaining weight fast.

**Obesity** is now a global epidemic, affecting over **650 million adults worldwide** (WHO, 2024). The culprits? A rapid rise in **high-calorie, nutrient-poor diets**, often rich in sugar, saturated fats, and refined carbohydrates.

As countries industrialize, fast food becomes more accessible and affordable gradually replacing traditional diets once rich in fibre, plant-based protein, and whole grains.

This dietary shift leads to an **energy imbalance**, causing excess body fat accumulation and elevating risks of **type 2 diabetes, heart disease, and cancer**.

> **How do changing intakes of sugar, fat, protein, and carbs shape the global obesity crisis?**

This project explores that question using global data from **1990 to 2022**, investigating the correlations between **macronutrient intake** and **obesity rates** over time and geography.


# Data Sources & Preprocessing

### Where Our Data Comes From

To explore the nutrition-obesity link, we combined **three open global datasets**:

- **Obesity Rates (WHO GHO)**  
  > Age-standardized % of obese adults (BMI ≥ 30), 1990–2022  

- **Diet Composition (FAO)**  
  > Per capita daily kilocalories from food groups like sugar, fats, oils, etc.

- **Macronutrient Supply (Our World in Data)**  
  > Average kcal/day from carbohydrates, fat, and animal vs. plant protein

### Preprocessing Steps

- Merged datasets by `Country` and `Year`
- Selected variables: **Obesity Rate**, **Sugar**, **Fat**, **Carbs**, **Animal & Plant Protein**
- Filtered for complete cases (no missing kcal values)
- Final dataset: **5,600+ country-year observations (1990–2022)**

This cleaned dataset is the foundation of our interactive visual analysis.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
library(tidyr)
library(plotly)
library(readr)

# Load datasets
obesity_data <- read_csv("C:/Users/prath/Downloads/obesity-prevalence-adults-who-gho/obesity-prevalence-adults-who-gho.csv")
diet_data <- read_csv("C:/Users/prath/Downloads/dietary-composition-by-country/dietary-composition-by-country.csv")
calories_data <- read_csv("C:/Users/prath/Downloads/daily-caloric-supply-derived-from-carbohydrates-protein-and-fat/daily-caloric-supply-derived-from-carbohydrates-protein-and-fat.csv")

# Step 1: Clean Obesity Data
obesity_clean <- obesity_data %>%
  rename(
    Country = Entity,
    Obesity_Rate = `Prevalence of obesity among adults, BMI >= 30 (age-standardized estimate) (%) - Sex: both sexes - Age group: 18+  years of age`
  ) %>%
  select(Country, Year, Obesity_Rate)

# Step 2: Clean Diet Composition Data
diet_clean <- diet_data %>%
  rename(Country = Entity) %>%
  select(Country, Year,
         Sugar_kcal = `Sugar & Sweeteners | 00002909 || Food available for consumption | 0664pc || kilocalories per day per capita`)

# Step 3: Clean Macronutrient Composition Data
calories_clean <- calories_data %>%
  rename(Country = Entity) %>%
  select(Country, Year,
         Animal_Protein_kcal = `Daily calorie supply per person that comes from animal protein`,
         Plant_Protein_kcal = `Daily calorie supply per person that comes from vegetal protein`,
         Fat_kcal = `Daily calorie supply per person from fat`,
         Carbs_kcal = `Daily calorie supply per person from carbohydrates`)

# Step 4: Merge all data
merged_df <- obesity_clean %>%
  inner_join(diet_clean, by = c("Country", "Year")) %>%
  inner_join(calories_clean, by = c("Country", "Year"))

# Step 5: Filter and select clean_df
clean_df <- merged_df %>%
  filter(Year >= 1990) %>%
  select(Country, Year, Obesity_Rate, Sugar_kcal, Fat_kcal, Carbs_kcal, Animal_Protein_kcal, Plant_Protein_kcal) %>%
  filter(!is.na(Obesity_Rate), !is.na(Sugar_kcal), !is.na(Fat_kcal), !is.na(Carbs_kcal))

```

# Global Macronutrient Intake Trends (1990–2022)

## Are We Eating Differently Over Time?

This stacked area chart visualizes the average global intake (in kcal/day) of five macronutrients from 1990 to 2022:

### Key Observations:
- **Carbohydrates** remain the largest energy source, though their share has slightly declined.
- **Fat intake** has steadily increased, reflecting dietary shifts toward processed and high-fat foods.
- **Sugar consumption** rose modestly, especially post-2000.
- **Animal protein** has gradually overtaken 
- **plant protein** in global diets, especially in industrialized nations.

```{r}

global_macro <- clean_df %>%
  group_by(Year) %>%
  summarise(
    Sugar = mean(Sugar_kcal, na.rm = TRUE),
    Fat = mean(Fat_kcal, na.rm = TRUE),
    Carbs = mean(Carbs_kcal, na.rm = TRUE),
    Animal_Protein = mean(Animal_Protein_kcal, na.rm = TRUE),
    Plant_Protein = mean(Plant_Protein_kcal, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = -Year, names_to = "Macronutrient", values_to = "kcal")

# Define stacked order
macro_order <- c("Sugar", "Plant_Protein", "Animal_Protein", "Fat", "Carbs")
global_macro$Macronutrient <- factor(global_macro$Macronutrient, levels = macro_order)

# Create interactive plot
plot_ly(global_macro,
        x = ~Year, y = ~kcal, color = ~Macronutrient,
        colors = c("Carbs" = "#e07a5f", "Fat" = "#3366CC", "Animal_Protein" = "#9d4edd",
                   "Plant_Protein" = "#2E7D32", "Sugar" = "#f6c177"),
        type = 'scatter', mode = 'none', stackgroup = 'one', fill = 'tonexty',
        text = ~paste0("Year: ", Year, "<br>Macronutrient: ", Macronutrient,
                       "<br>Avg Intake: ", round(kcal), " kcal/day"),
        hoverinfo = 'text') %>%
  layout(title = "Global Average Macronutrient Intake (kcal/day) (1990–2022)",
         xaxis = list(title = "Year"),
         yaxis = list(title = "kcal/day"),
         legend = list(title = list(text = "Macronutrient")))


```

### Implication:
These macronutrient shifts reflect a **nutrition transition** from traditional diets to modern, energy-dense foods that aligns with rising obesity rates.

> Let’s see how this connects with the global obesity curve.




# **Global Obesity and Nutrition Trends (1990–2022)**

#### *A Rising Curve with a Deeper Cause*

Over the last three decades, something has quietly but dramatically shifted across the world.

In 1990, global obesity affected fewer than 1 in 10 adults. By 2022, that number had surged past 1 in 5, and it continues to rise.

What changed?

As countries urbanized and food systems evolved, traditional diets gave way to modern, energy-dense foods. These new diets are rich in added sugars, processed fats, and animal-based products, while often lacking essential fiber and nutrients.


```{r plot_global_obesity_trend, echo=FALSE, message=FALSE, warning=FALSE}

# Load plotly if not already
library(plotly)

# Ensure global_trend is created first
global_trend <- clean_df %>%
  group_by(Year) %>%
  summarise(Global_Obesity = mean(Obesity_Rate, na.rm = TRUE))

# Plotly chart with two traces
p <- plot_ly()

# Add trend line
p <- add_trace(p,
               data = global_trend,
               x = ~Year, y = ~Global_Obesity,
               type = 'scatter', mode = 'lines',
               line = list(color = 'black'),
               name = 'Trend Line (Black)')

# Add data points
p <- add_trace(p,
               data = global_trend,
               x = ~Year, y = ~Global_Obesity,
               type = 'scatter', mode = 'markers',
               marker = list(color = 'blue', size = 6),
               name = 'Data Points (Blue Dots)',
               text = ~paste("Year:", Year,
                             "<br>Obesity Rate:", round(Global_Obesity, 2), "%"))

# Layout
p <- layout(p,
            title = "Global Average Obesity Rate Over Time (1990–2022)",
            xaxis = list(title = "Year"),
            yaxis = list(title = "Obesity Rate (%)"),
            legend = list(
              x = 0.02,
              y = 0.98,
              bgcolor = "rgba(255,255,255,0.5)",
              bordercolor = "black",
              borderwidth = 1
            ),
            hovermode = "x unified")

p



```
At the same time, physical activity declined and sedentary lifestyles became the norm. But the real story is captured in the data:

* The **black line** shows the consistent rise in global obesity prevalence from 1990 to 2022.
* The **blue dots** represent measured yearly averages, highlighting a clear and continuous upward trend.

This chart is not just data—it reflects a global shift in the way we eat, live, and move.

- The **global obesity rate** has more than **doubled**, climbing from ~10% in 1990 to **over 23%** by 2022.
- The trend aligns with increased **energy-dense, nutrient-poor diets**.
-  The sharpest growth occurred **post-2000**, coinciding with fast food proliferation and lifestyle changes.

> This global rise signals a deeper nutritional shift.

Up next: What **macronutrients** are most strongly linked to this pattern?



# What Drives Obesity? Macronutrient Correlations

## Identifying Key Dietary Drivers

To investigate potential contributors to the global rise in obesity, we examined correlations between macronutrient intake and obesity rates from 1990 to 2022.

### Key Insights:

- **Sugar** and **fat intake** are strongly and positively correlated with obesity rates.
- **Animal protein** shows moderate positive correlation.
- **Carbohydrates** have a weak or slightly negative correlation.
- **Plant protein** exhibits a weak negative association.


```{r}
cor_df <- clean_df %>%
  select(Obesity_Rate, Sugar_kcal, Fat_kcal, Carbs_kcal, Animal_Protein_kcal, Plant_Protein_kcal)

cor_matrix <- cor(cor_df, use = "complete.obs")

plot_ly(
  x = colnames(cor_matrix),
  y = rownames(cor_matrix),
  z = cor_matrix,
  type = "heatmap",
  colorscale = "RdBu", zmin = -1, zmax = 1, reversescale = TRUE
) %>%
  layout(title = "Correlation Between Obesity and Macronutrients",
         xaxis = list(tickangle = 45), yaxis = list(title = ""))


```
### Interpretation:

Diets high in sugar and fat appear to be key contributors to rising obesity rates, while plant-based diets may offer a protective effect.

This evidence highlights the nutritional transition driving global obesity trends.

> Next: We explore how each macronutrient individually relates to obesity rates.


# Obesity vs. Macronutrient Intake

### How Does Each Nutrient Influence Obesity?

This section explores how specific macronutrients correlate with obesity rates based on global data (1990–2022):

- **Sugar:** Strong positive correlation countries with higher sugar intake generally have higher obesity rates.
- **Fat:** Also positively associated with obesity, though less consistent than sugar.
- **Carbohydrates:** Weak or no clear association some high-carb diets correspond with lower obesity rates.
- **Protein:**
- *Animal protein* shows a mild positive trend.
- *Plant protein* tends to have a neutral or slightly protective effect.




```{r obesity_vs_macros_interactive, echo=FALSE, message=FALSE, warning=FALSE}


# ---------- 1. Create the five ggplots ----------
plots <- list(
  Sugar = ggplot(clean_df, aes(Sugar_kcal, Obesity_Rate)) +
    geom_point(alpha = 0.3, colour = "#1f77b4") +
    geom_smooth(method = "lm", colour = "black") +
    labs(title = "Obesity vs Sugar", x = "Sugar (kcal/day)", y = "Obesity Rate (%)"),

  Fat = ggplot(clean_df, aes(Fat_kcal, Obesity_Rate)) +
    geom_point(alpha = 0.3, colour = "#ff7f0e") +
    geom_smooth(method = "lm", colour = "black") +
    labs(title = "Obesity vs Fat", x = "Fat (kcal/day)", y = "Obesity Rate (%)"),

  Carbs = ggplot(clean_df, aes(Carbs_kcal, Obesity_Rate)) +
    geom_point(alpha = 0.3, colour = "#2ca02c") +
    geom_smooth(method = "lm", colour = "black") +
    labs(title = "Obesity vs Carbohydrates", x = "Carbs (kcal/day)", y = "Obesity Rate (%)"),

  `Animal Protein` = ggplot(clean_df, aes(Animal_Protein_kcal, Obesity_Rate)) +
    geom_point(alpha = 0.3, colour = "#d62728") +
    geom_smooth(method = "lm", colour = "black") +
    labs(title = "Obesity vs Animal Protein", x = "Animal Protein (kcal/day)", y = "Obesity Rate (%)"),

  `Plant Protein` = ggplot(clean_df, aes(Plant_Protein_kcal, Obesity_Rate)) +
    geom_point(alpha = 0.3, colour = "#9467bd") +
    geom_smooth(method = "lm", colour = "black") +
    labs(title = "Obesity vs Plant Protein", x = "Plant Protein (kcal/day)", y = "Obesity Rate (%)")
)

# ---------- 2. Convert to plotly & gather traces ----------
plt_list   <- lapply(plots, ggplotly, tooltip = "text")
trace_list <- do.call(c, lapply(plt_list, function(p) p$x$data))

grp_vec <- rep(names(plots), sapply(plt_list, function(p) length(p$x$data)))
vis_vec <- grp_vec == names(plots)[1]  # default visible: Sugar

# ---------- 3. Build the switchable figure ----------
p <- plot_ly()
for (i in seq_along(trace_list)) {
  tr <- trace_list[[i]]
  p  <- add_trace(p,
    x = tr$x, y = tr$y,
    type = tr$type, mode = tr$mode,
    marker = tr$marker, line = tr$line,
    name = grp_vec[i],
    visible = vis_vec[i],
    hoverinfo = tr$hoverinfo %||% "x+y"
  )
}

# ---------- 4. Add dropdown menu ----------
buttons <- lapply(unique(grp_vec), function(grp) {
  list(method = "restyle",
       args = list("visible", grp_vec == grp),
       label = grp)
})

# ---------- 5. Add annotations ----------
explain_txt <- paste(
  "<b>Legend guide</b><br>",
  "● Points = country-year observations<br>",
  "― Solid line = linear regression fit<br>",
  "▬ Faint band = 95% confidence interval"
)

p <- layout(p,
  title = "Obesity vs Macronutrient Intake (Interactive)",
  xaxis = list(title = "Macronutrient intake (kcal/day)"),
  yaxis = list(title = "Obesity Rate (%)"),
  updatemenus = list(
    list(type = "dropdown", direction = "down",
         x = 0, y = 1.15, buttons = buttons,
         showactive = TRUE)
  ),
  annotations = list(
    list(text  = explain_txt,
         xref = "paper", yref = "paper",
         x = 1.02, y = 1,
         showarrow = FALSE, align = "left",
         font = list(size = 11))
  ),
  margin = list(r = 140)
)

p


```


## Summary: What Drives Obesity?

- Obesity has more than doubled globally since 1990.
- Sugar and fat intake show strong positive relationships with obesity.
- Plant-based diets appear to offer modest protective effects.
- The transition to energy-dense, processed foods is a major global concern.

> These patterns offer a big-picture view, but which regions are most affected?



# Country Focus: Top 10 by Average Obesity Rate (1990–2022)

- Small island nations in the **Pacific region** dominate the list.
- These countries have undergone rapid **dietary westernization**, often replacing traditional, nutrient-dense diets with imported processed foods.
- **Tonga, Nauru, and Tuvalu** consistently show average obesity rates above **60%**.
- High prevalence is linked to affordability of processed imports and limited access to fresh, local foods.

> Global solutions must consider **localized dietary shifts** and **cultural influences**.


```{r}
top10 <- clean_df %>%
  group_by(Country) %>%
  summarise(Avg_Obesity = mean(Obesity_Rate, na.rm = TRUE)) %>%
  arrange(desc(Avg_Obesity)) %>%
  slice(1:10)

plot_ly(top10,
        x = ~Avg_Obesity, y = ~reorder(Country, Avg_Obesity),
        type = 'bar', orientation = 'h',
        marker = list(color = '#f6c177'),
        text = ~paste0(Country, ": ", round(Avg_Obesity, 1), "%"),
        hoverinfo = 'text') %>%
  layout(title = "Top 10 Countries by Average Obesity Rate",
         xaxis = list(title = "Obesity Rate (%)"),
         yaxis = list(title = "Country"))


```



# Can We Predict Obesity From Diet?

- A **linear regression** shows that sugar intake alone explains **~29%** of the variance in obesity.
- A **multiple regression model** using sugar, fat, carbs, and proteins improves explanatory power to **~37% (R² = 0.37)**.


```{r, include=FALSE }
# Linear Model: Obesity vs Sugar
model_sugar <- lm(Obesity_Rate ~ Sugar_kcal, data = clean_df)
summary(model_sugar)

# Multiple Regression Model
model_multi <- lm(Obesity_Rate ~ Sugar_kcal + Fat_kcal + Carbs_kcal +
                    Animal_Protein_kcal + Plant_Protein_kcal, data = clean_df)
summary(model_multi)
```

```{r, echo=FALSE }
# Predicted vs Actual
predicted <- predict(model_multi, clean_df)

p_pred <- ggplot(data.frame(Actual = clean_df$Obesity_Rate, Predicted = predicted),
                 aes(x = Predicted, y = Actual)) +
  geom_point(alpha = 0.3, color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Predicted vs Actual Obesity Rate", x = "Predicted", y = "Actual")

ggplotly(p_pred)

```


### What the Prediction Plot Shows:
- The red line represents perfect prediction (Actual = Predicted).
- The scatter shows some variability indicating **other lifestyle, economic, and genetic factors** also play a role.
- Still, **macronutrient intake** has a measurable and statistically significant effect on obesity trends.



# Conclusion

Over the past three decades, global obesity rates have **more than doubled**, rising alongside significant shifts in dietary patterns.
This analysis uncovered clear patterns:

- **High intake** of sugar, fat, and animal protein is **strongly linked** to higher obesity rates.  
- Nations most affected including many **Pacific islands** and **resource-rich countries** are experiencing rapid **nutrition and lifestyle transitions**.  
- In contrast, **plant-based diets** and **moderate sugar consumption** appear to offer **protective benefits**.
Together, these findings paint a powerful picture:  
**Obesity is not just a personal health issue it's a global nutritional consequence.**

> Effective policy, education, and access to healthier foods are essential to reverse this rising tide.

## Reference

[1]Ritchie, H., & Roser, M. (2024). Obesity prevalence among adults (WHO). Our World in Data. https://ourworldindata.org/grapher/obesity-prevalence-adults-who-gho

[2]Ritchie, H., & Roser, M. (2024). Daily caloric supply from carbohydrates, protein and fat. Our World in Data. https://ourworldindata.org/grapher/daily-caloric-supply-derived-from-carbohydrates-protein-and-fat

[3]Ritchie, H., & Roser, M. (2024). Dietary composition by country. Our World in Data. https://ourworldindata.org/grapher/dietary-composition-by-country
